---
output:
  github_document:
    df_print: tibble
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  tidy = TRUE,
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# uktrade

<!-- badges: start -->
[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

The goal of `uktrade` is to provide convenient wrapper functions for loading HMRC data in R.

## Installation

You can install the latest version of `uktrade` from this GitHub repository using `remotes`:

``` r
remotes::install_github("pvdmeulen/uktrade")
```
## HMRC data

Her Majesty's Revenue & Customs (HMRC) is the :uk: UK's customs authority. Data on UK trade is available on their [uktradeinfo.com](https://www.uktradeinfo.com/) website, which is collected through a combination of customs declarations and Intrastat surveys (in the case of [some EU trade](https://www.gov.uk/intrastat)).

In the beginning of 2021, the [uktradeinfo.com](https://www.uktradeinfo.com/) website was updated. This update also introduced an Application Programming Interface (API) for accessing bulk trade (and trader) data. For smaller trade data extracts (< 30,000 rows), the [online tool](https://www.uktradeinfo.com/trade-data/) may be sufficient.

## This package

This package contains four functions:

|Function         |Description                                            |Status                                         |
|-----------------|-------------------------------------------------------|-----------------------------------------------|
|`load_ots()`     |a function for loading Overseas Trade Statistics data  |:yellow_circle: experimental (use at own risk) |
|`load_rts()`     |a function for loading Regional Trade Statistics data  |:yellow_circle: experimental (use at own risk) |
|`load_trader()`  |a function for loading trader data                     |:red_circle: *planned*                         |
|`load_custom()`  |a function for loading a custom URL                    |:yellow_circle: experimental (use at own risk) |

All of these functions will output a `dataframe` object (or `tibble`) with the desired data, and are able to keep track of paginated results (in batches of 30,000 rows) as well as the API request limit of 60 requests per minute.

The first three are convenient wrapper functions which should make loading basic datasets easier (with the assumption that further data manipulation will be done in R after loading the data). Of course, the API allows for extensive customisation in what data is obtained. For this purpose, `load_custom()` allows you to specify a custom URL instead. This allows you to be more specific in your request (and in the case of trader data, expand specific columns to get more information).

## Example using OTS and RTS data

These functions are convenient wrappers for loading trade data - either UK trade data using OTS, or regional UK trade data using RTS. These functions will load the raw data and optionally join results with lookups obtained from the API (using, for example, the /Commodity and /Country endpoints). This makes data easy to read by humans, but also larger (more columns containing text).

### OTS

Loading all UK trade of single malt Scotch whisky (CN8 code 22083030) and bottled gin (22085011) for 2019 is done like so:

```{r load_ots example}
library(uktrade)
data <- load_ots(month = 201901:201912, commodity = c(22083030, 22085011), join_lookup = FALSE, output = "tibble")

# Results are now in a tibble (set output to "df" to obtain a dataframe):
data
```

As you can see, results are not easily interpretable as they stand. The HMRC API also has lookups which can be loaded separately using `load_custom()`, or joined automatically:

```{r load_ots example with join}
library(uktrade)

# Load specific lookups separately:
commodity_lookup <- load_custom(endpoint = "Commodity")

# Or join automatically with the `join_lookup = TRUE` option:
data <- load_ots(month = 201901:201912, commodity = c(22083030, 22085011), join_lookup = TRUE)

data
```

Note that HMRC's API isn't working entirely as expected at the moment. At the moment, loading aggregate data (such as all spirits, HS4 code 2208) returns an empty dataframe.

```{r load_ots example with join and HS4}
library(uktrade)
data <- load_ots(month = 201901:201912, commodity = 2208, join_lookup = FALSE)

data
```

However, specifying `commodity = 0` will load all commodities aggregated (specifying `NULL` will load all detailed commodities and may take considerable time):

```{r load_ots example with join and commodity0}
library(uktrade)
data <- load_ots(month = 201901, commodity = 0, join_lookup = FALSE)

data
```

### RTS

Loading all UK regional trade of 2-digit SITC Division '00 - Live animals other than animals of division 03' for 2019 is done like so:

```{r load_rts example}
library(uktrade)
data <- load_rts(month = 201901:201912, sitc = 00, join_lookup = TRUE)

data
```

## Example using trader data

Trader code is a work in progress...

## Example using a custom URL

This example uses the `load_custom()` function to replicate the example given in the [API documentation](https://www.uktradeinfo.com/api-documentation/) for finding all traders that exported ‘Live horses (excl. pure-bred for breeding)’ from the ‘CB8’ post code in 2019. This way of loading data is possible, but will require more data manipulation after obtaining the results. This function is also the workhorse (no pun intended) behind the other functions.

```{r load_custom example}
library(uktrade)
data <- load_custom(endpoint = "Commodity", custom_search = "?$filter=Hs6Code eq '010129'&$expand=Exports($filter=MonthId ge 201901 and MonthId le 201912 and startswith(Trader/PostCode, 'CB8'); $expand=Trader)", output = "tibble")

data
```

Note that the expanded columns, Exports and Trader, both need to be expanded as they are contained in a nested <list> column. When unnested using `tidyr::unnest()`, we can see the final results.
The second expanded column (Trader) itself contains 8 columns (which are TraderId, CompanyName, five Address columns, and PostCode).

```{r unnesting columns}
library(tidyr)
tidyr::unnest(data, Exports, names_repair = "unique")
```

## MIT License

You're free to use this code and its associated documentation without restriction. This includes both personal and commercial use. However, there is no warranty provided or any liability on my part. See the LICENSE file for more information.
